### 2022-10-04 chapters/literature/draft_citations.tex
add : summary for programmming langaugess

We will place a greater emphasis on the software level to decrease the amount of energy that is used, more particularly in the execution phase of the program cycle.
We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
To do this, we will elaborate on a benchmarking process and a set of tools intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
Thus, we will begin by examining state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field. Then, we will narrow these practices down to computer science so we can finally adapt them to energy consumption.

@@

\subsection{Hardware Tools}
Nowadays, most high-performance computing systems (HPC) implement a tool to report the nodes' energy consumption for monitoring and administration.
Those tools are mainly integrated within the power supply units (PSU) or the power distribution units (PDU).
Then, they provide an interface and a log to follow the energy consumption history.
Despite their scalability and ease of integration, such tools lack both spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the time, they have a shallow sampling frequency.
Most of those tools are provided directly by the manufacturers.
Such as \em{IBM EnergyScale technology}~\cite{mccreary2007energyscale,caldeira2014ibm,caldeiraibm} or \em{Dell poweredge}~\cite{lovicott2009thermal}, MEGware Cluststafe~\cite{breitbart2015case}.
As said earlier, the true purpose of those tools is more monitoring than analyzing energy consumption.
%% add something to link with wattsup pro 
WattsUp Pro is a device that can be installed between the power source of the machine and the system under test.
It allows a sampling frequency of up to 1\,Hz. It has an internal memory to store a wide variety of data, such as the maximum voltage and current, that can later be exported via USB port for personal usage or lined to graph programs like Logger Pro or LabQuest.
The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test~\cite{hirst2013watts}
Despite its high temporal sampling, WattsUp Pro lacks spatial granularity since it monitors the energy consumption of the whole system.
To have finer granularity, we need to isolate the energy consumption of each component.
% For this, we will need more intrusive tools such.

@@

Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
We can observe a consedirable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.

@@

We start with the work of \citeauthor{pereira_energy_2017} where they did an energy consumption comparison analysis of the most popular programming languages.
The paper recommends combining some of these languages to enhance code quality while considering execution time, memory utilization, and energy consumption using the Pareto optimum~\cite{hochman1969pareto}. Some of the study's findings demonstrate that interpreted languages, such as Python, have lower energy efficiency than compiled languages, such as C or Rust. The research also offers language combinations that developers might use to improve energy economy, execution speed, and memory utilization.
This paper's findings were based on the game benchmark, the most famous benchmark comparing several programming languages.

This paper defines a ranking of energy efficiency in programming languages. We considered a set of computing problems implemented in ten well-known programming languages and monitored the energy consumed when executing each language. Our preliminary results show that although the fastest languages tend to be the lowest energy-consuming ones, there are other interesting cases where slower languages are more energy-efficient than faster ones.

@@

\section{Conclusion}
As we have seen in state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
Furthermore, the execution phase took the attention of many researchers because it is the part where the most energy is consumed.
This thesis will focus on that aspect as well.
However, unlike most work done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language to how to tune some features of a framework to make the software consumes less energy.
To do so, we use the empirical approach due to its consistency for the moment.
Unlike the performance, which is essentially related to the algorithm's complexity, the energy consumption is more impacted by the hardware.
Therefore, to optimize the energy consumption, we choose a spiral method based on 3 phases:
\begin{enumerate}
    \item executes the code,
    \item measure the program,
    \item infers the guidelines.
\end{enumerate}

>>>>

We will place a greater emphasis on the software level to decrease the amount of energy that is used, more particularly in the execution phase of the program cycle.
We will be proceeding through an empirical analysis of the energy consumption of the software while changing some components of the source code without impacting its behavior.
To do this, we will elaborate on a benchmarking process and a set of tools intended to assist practitioners in better comprehending and optimizing the energy usage of their applications.
Thus, we will begin by examining the state of the art in empirical analysis and retrieving the best empirical experimentation methodologies in the research field. Then, we will narrow these practices down to computer science so we can finally adapt them to energy consumption.

@@

\subsection{Hardware Tools}
Nowadays, most high-performance computing systems (HPC) implement a tool to report the nodes' energy consumption for monitoring and administration.
Those tools are mainly integrated within the power supply units (PSU) or the power distribution units (PDU).
Then, they provide an interface and a log to follow the energy consumption history.
Despite their scalability and ease of integration, such tools lack both spatial and temporal granularity since they monitor the whole energy of the nodes, and most of the time, they have a shallow sampling frequency.
Most of those tools are provided directly by the manufacturers.
Such as \em{IBM EnergyScale technology}~\cite{mccreary2007energyscale,caldeira2014ibm,caldeiraibm} or \em{Dell poweredge}~\cite{lovicott2009thermal}, MEGware Cluststafe~\cite{breitbart2015case}.
As said earlier, the true purpose of those tools is more monitoring than analyzing energy consumption.
%% add something to link with wattsup pro 
WattsUp Pro is a device that can be installed between the power source of the machine and the system under test.
It allows a sampling frequency of up to 1\,Hz. It has an internal memory to store a wide variety of data, such as the maximum voltage and current, that can later be exported via a USB port for personal usage or lined to graph programs like Logger Pro or LabQuest.
The main advantage of this tool is the ability to monitor the energy consumption from a different device which will reduce the risk of interference with the energy consumption of the test~\cite{hirst2013watts}
Despite its high temporal sampling, WattsUp Pro lacks spatial granularity since it monitors the energy consumption of the whole system.
To have finer granularity, we need to isolate the energy consumption of each component.
% For this, we will need more intrusive tools such.

@@

Figure~\ref{fig:motivation} illustrates this variation problem as a violin plot of $20$ executions of the benchmark \emph{Conjugate Gradient} (\textsf{CG}) taken from the \emph{NAS Parallel Benchmarks} (NBP) suite~\cite{Bailey:1991:NPB:125826.125925}, on $4$ nodes of an homogeneous cluster (the cluster \textsf{Dahu} described in Table~\ref{table:g5k}) at 50\,\% workload.
We can observe a considerable variation of the energy consumption, not only among homogeneous nodes but also at the scale of a single node, reaching up to $25\,\%$ in this example.

@@

We start with the work of \citeauthor{pereira_energy_2017} where they did an energy consumption comparison analysis of the most popular programming languages.
The paper recommends combining some of these languages to enhance code quality while considering execution time, memory utilization, and energy consumption using the Pareto optimum~\cite{hochman1969pareto}. Some of the study's findings demonstrate that interpreted languages, such as Python, have lower energy efficiency than compiled languages, such as C or Rust. The research also offers language combinations that developers might use to improve energy economy, execution speed, and memory utilization.
This paper's findings were based on the game benchmark\footnote{\url{https://benchmarksgame-team.pages.debian.net/benchmarksgame/index.html}}, the most famous benchmark comparing several programming languages.

@@

\section{Conclusion}
As we have seen in the state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
Furthermore, the execution phase took the attention of many researchers because it is the part where the most energy is consumed.
This thesis will focus on that aspect as well.
However, unlike most work done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language to how to tune some features of a framework to make the software consumes less energy.
To do so, we use the empirical approach due to its consistency for the moment.
Unlike the performance, which is essentially related to the algorithm's complexity, the energy consumption is more impacted by the hardware.
Therefore, to optimize the energy consumption, we choose a spiral method based on 3 phases:
\begin{enumerate}
    \item executes the code,
    \item measure the program,
    \item infers the guidelines.
\end{enumerate}

### 2022-10-04 chapters/programming_languages/programming_languages.tex
add : example
add : summary for programmming langaugess

\section{Investigating Web Application Frameworks}\label{sec:webframework}
\import{\currfiledir/web_frameworks}{web_frameworks}

\section{Conclusion \note{missing}}

>>>>

\section{Investigating Web Application Frameworks}\label{sec:webframework}

\import{\currfiledir/web_frameworks}{web_frameworks}

\section{Summary}

this chapter aims to extend the work of \citeauthor{pereira_energy_2017}~\cite{pereira_energy_2017} by studying the energy footprint on other aspects than micro-benchmarking.
The main focus of this chapter is the behavior of programming languages within the context of web servers. To do so, we have chosen to study cases.
The first experiment is the impact of programming languages when handling the RPC protocol. We used the official implementation of \textsc{gRPC} library to build 25 frameworks using 17 programming languages.
In the second experiment, we look at how web frameworks affect energy consumption by creating a basic web application and comparing its performance, latency, and energy usage using the most common web frameworks.
Overall we had $261$ implementation using $35$ and three different databases.
For each implementation, we considered seven typical scenarios. Each scenario is tested with several stress levels to study the scalability of the web frameworks. Therefore the final results are an $8,750$ use case.

Our studies revealed the absence of a clear winner in terms of energy efficiency. However, when we take into consideration the cost of a single request. C++ and Rust were the top-ranked. These results confrim the work done by \citeauthor{pereira_energy_2017}. On the other hand, Unlike the results shown in the paper \cite{pereira_energy_2017}, \textsf{PHP} exhibits a higher performance while keeping an average power consumption 
This result is due to the nature of \textsf{PHP} language, which is dedicated to the web and not general-purpose programming. 
While most of the frameworks consumed around $100 watts$, \textsf{JAVA} based frameworks consumed around $220 watts$. which made it one of the most energy-consuming in the long run, especially when considering the times when there is low traffic.

Finally, this study revealed that each framework had different behaviors toward scalability. Some of them increased the average power consumption while trying to keep up with the performance. Others, on the other hand, could keep the average power consumption while decreasing the performance. Furthermore, there was a case where there was a mix between both strategies.

As there is no final answer to the question of which framework is the most energy-efficient, we provide developers with a tool to help them compare between frameworks based on their needs. The criteria we used to compare frameworks are the scenario type, the stress level, and the database type. The tool is available at \url{https://github.com/chakib-belgaid/greenboard}
